---
title: "Data"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, results='asis', message=FALSE, warning=FALSE, collapse=TRUE)
```

```{r}
library(heatmetrics)
ls("package:heatmetrics") # functions in the heatmetrics package
```

Data source:

Muñoz Sabater, J. (2019): ERA5-Land hourly data from 1950 to present. Copernicus Climate Change Service (C3S) Climate Data Store (CDS). DOI: 10.24381/cds.e2161bac (Accessed on 27-01-2026)

Hersbach, H., Bell, B., Berrisford, P., Biavati, G., Horányi, A., Muñoz Sabater, J., Nicolas, J., Peubey, C., Radu, R., Rozum, I., Schepers, D., Simmons, A., Soci, C., Dee, D., Thépaut, J-N. (2023): ERA5 hourly data on single levels from 1940 to present. Copernicus Climate Change Service (C3S) Climate Data Store (CDS), DOI: 10.24381/cds.adbb2d47 (Accessed on 27-01-2026)


```{r}
library(ncdf4)
era5land <- nc_open("era5-land.nc") # era5-land variables (doesn't have fdir)
names(era5land$var)

era5 <- nc_open("era5.nc") # era5 fdir variable
names(era5$var)
```



