---
title: "preliminary analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(terra)
library(stringr)
library(dplyr)
library(ggplot2)
library(zoo)
```


## temporal analysis

```{r}
# load all files

files <- list.files(
  "wbgt",
  pattern = "\\.nc$",
  full.names = TRUE
)

df <- data.frame(
  file = files
)

df$year <- str_extract(basename(files), "\\d{4}") |> as.integer()

df$month <- str_extract(
  basename(files),
  "(?<=_)\\d{2}(?=\\.nc)"
) |> as.integer()
```


monthly min, mean, max
```{r}
# create monthly min, mean, max

monthly_stats <- lapply(seq_along(files), function(i){

  r <- rast(files[i], subds = "wbgt")

  vals <- values(r)

  data.frame(
    year = df$year[i],
    month = df$month[i],
    min_wbgt = min(vals, na.rm=TRUE),
    mean_wbgt = mean(vals, na.rm=TRUE),
    max_wbgt = max(vals, na.rm=TRUE),
    pct_over_32 = mean(vals > 32, na.rm=TRUE) * 100  # percent of CA area that is > 32C (extreme)
  )
})

monthly_stats <- bind_rows(monthly_stats)
```

Monthly, annual, and seasonal min, mean, max
```{r}
# monthly time series

monthly_stats <- monthly_stats %>%
  mutate(date = as.yearmon(paste(year, month), "%Y %m"))

ggplot(monthly_stats, aes(x = date, y = mean_wbgt)) + geom_line()


# annual time series
annual <- monthly_stats %>%
  group_by(year) %>%
  summarize(
    min_wbgt = min(min_wbgt),
    mean_wbgt = mean(mean_wbgt),
    max_wbgt = max(max_wbgt),
    pct_over_32 = mean(pct_over_32)
  )

ggplot(annual, aes(year, mean_wbgt)) + geom_line()


# seasonal mean
monthly_stats <- monthly_stats %>%
  mutate(
    season = case_when(
      month %in% c(12,1,2) ~ "winter",
      month %in% c(3,4,5) ~ "spring",
      month %in% c(6,7,8) ~ "summer",
      month %in% c(9,10,11) ~ "autumn"
    ))

seasonal <- monthly_stats %>%
  group_by(year, season) %>%
  summarize(
    mean_wbgt = mean(mean_wbgt),.groups = "drop")

ggplot(seasonal, aes(x = year, y = mean_wbgt, color = season)) + geom_line() + theme_minimal()
ggplot(seasonal, aes(season, mean_wbgt)) + geom_boxplot()


# aggregated monthly mean
monthly <- monthly_stats %>%
  group_by(month) %>%
  summarize(
    mean_wbgt = mean(mean_wbgt)
  )

ggplot(monthly, aes(month, mean_wbgt)) + geom_line()

# another one split by year or groups of years
ggplot(monthly_stats, aes(x = month, y = mean_wbgt, group = year, color = factor(year))) +
  geom_line() + labs(x = "Month", y = "Mean WBGT", color = "Year") + theme_minimal()
```

extreme WBGT
```{r}
# percent of California area that has WBGT > 32
ggplot(annual, aes(year, pct_over_32)) + geom_line() + 
  labs(y = "percent of California area that has WBGT > 32")
```


Annual mean WBGT trend map
```{r}
years <- str_extract(basename(files), "\\d{4}") |> as.integer()
months <- str_extract(
  basename(files),
  "(?<=_)\\d{2}(?=\\.nc)"
) |> as.integer()

# annual mean WBGT raster
unique_years <- sort(unique(years))

annual_mean_list <- list()

for(y in unique_years){

  yr_files <- files[years == y]

  r <- rast(yr_files)

  # annual mean across all timesteps. Mean WBGT for that year at every grid cell.
  annual_mean <- app(r, mean, na.rm=TRUE)

  annual_mean_list[[as.character(y)]] <- annual_mean
}

# stack annual rasters
annual_stack <- rast(annual_mean_list)  # Build a time series per pixel
names(annual_stack) <- unique_years

# compute trend per gridcell
trend_fun <- function(x){

  if(all(is.na(x))) return(NA)

  yrs <- unique_years

  fit <- lm(x ~ yrs) # Fit a linear regression at each grid cell

  slope <- coef(fit)[2]

  # convert to °C/year
  slope
}

trend_map <- app(
  annual_stack,
  trend_fun
)

# plot trend map. Each pixel contains change in annual mean WBGT per year
# Example interpretation: +0.35 °C/year means: Annual mean WBGT increased by 0.35°C every 10 years at that location.

ca_shape <- vect("deriving urban/ca_grid/ERA5_Land_grid_CAfull.shp")
# plot raster cropped to CA
trend_ca <- mask(crop(trend_map, ca_shape), ca_shape)
plot(trend_ca, main = "Annual Mean WBGT Trend (°C/year)")
```

Extreme WBGT trend map
```{r}

```



To do: grid-cell/regional level trends
map of extreme WBGT across CA, comparisons across major regions (coastal, inland, valley, desert, mountains), identification of WBGT hotspots and areas with the largest increases over time

