---
title: "preliminary analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(terra)
library(stringr)
library(dplyr)
library(ggplot2)
library(zoo)
library(purrr)
```


## temporal analysis

```{r}
# load all files

files <- list.files(
  "wbgt",
  pattern = "\\.nc$",
  full.names = TRUE
)

df <- data.frame(
  file = files
)

df$year <- str_extract(basename(files), "\\d{4}") |> as.integer()

df$month <- str_extract(
  basename(files),
  "(?<=_)\\d{2}(?=\\.nc)"
) |> as.integer()
```


crop data to CA shape
```{r}
# read CA shape
ca_shape <- vect("deriving urban/ca_grid/ERA5_Land_grid_CAfull.shp")

# output directory
dir.create("wbgt_ca", showWarnings = FALSE)

# loop through files
for(i in seq_len(nrow(df))) {

  f <- df$file[i]

  cat("Processing:", basename(f), "\n")

  # read raster
  r <- rast(f)

  # project shape if needed
  ca_proj <- project(ca_shape, crs(r))

  # crop first (faster)
  r_crop <- crop(r, ca_proj)

  # mask to exact CA boundary
  r_ca <- mask(r_crop, ca_proj)

  # output filename
  out_file <- file.path(
    "wbgt_ca",
    paste0(tools::file_path_sans_ext(basename(f)), "_CA.nc")
  )

  # write NetCDF
  writeCDF(
    r_ca,
    filename = out_file,
    overwrite = TRUE
  )
}
```


open cropped WBGT files
```{r}
files_ca <- list.files(
  "wbgt_ca",
  pattern = "\\.nc$",
  full.names = TRUE
)

df_ca <- data.frame(
  file = files_ca
)

df_ca$year <- str_extract(basename(files_ca), "\\d{4}") |> as.integer()

df_ca$month <- str_extract(
  basename(files_ca),
  "(?<=_)\\d{2}(?=_CA\\.nc)"
) |> as.integer()
```


monthly min, mean, max
```{r}
# create monthly min, mean, max

monthly_stats <- lapply(seq_along(files_ca), function(i){

  r <- rast(files_ca[i], subds = "wbgt")

  vals <- values(r)

  data.frame(
    year = df_ca$year[i],
    month = df_ca$month[i],
    min_wbgt = min(vals, na.rm=TRUE),
    mean_wbgt = mean(vals, na.rm=TRUE),
    max_wbgt = max(vals, na.rm=TRUE),
    pct_over_32 = mean(vals > 32, na.rm=TRUE) * 100  # percent of CA area that is > 32C (extreme)
  )
})

monthly_stats <- bind_rows(monthly_stats)
```

Monthly, annual, and seasonal min, mean, max
```{r}
# monthly time series --- add min and max lines

monthly_stats <- monthly_stats %>%
  mutate(date = as.yearmon(paste(year, month), "%Y %m"))

ggplot(monthly_stats, aes(x = date, y = mean_wbgt)) + geom_line()

ggplot(monthly_stats, aes(x = date)) + geom_line(aes(y = mean_wbgt, color = "Mean")) +
  geom_line(aes(y = min_wbgt, color = "Min")) + geom_line(aes(y = max_wbgt, color = "Max")) +
  labs(x = "Date", y = "WBGT", color = "Statistic") + theme_minimal()

# annual time series
annual <- monthly_stats %>%
  group_by(year) %>%
  summarize(
    min_wbgt = min(min_wbgt),
    mean_wbgt = mean(mean_wbgt),
    max_wbgt = max(max_wbgt),
    pct_over_32 = mean(pct_over_32)
  )

ggplot(annual, aes(year, mean_wbgt)) + geom_line()


# annual time series split by seasonal
monthly_stats <- monthly_stats %>%
  mutate(
    season = case_when(
      month %in% c(12,1,2) ~ "winter",
      month %in% c(3,4,5) ~ "spring",
      month %in% c(6,7,8) ~ "summer",
      month %in% c(9,10,11) ~ "autumn"
    ))

seasonal <- monthly_stats %>%
  group_by(year, season) %>%
  summarize(
    mean_wbgt = mean(mean_wbgt),.groups = "drop")

ggplot(seasonal, aes(x = year, y = mean_wbgt, color = season)) + geom_line() + theme_minimal()
ggplot(seasonal, aes(season, mean_wbgt)) + geom_boxplot()


# aggregated monthly mean
monthly <- monthly_stats %>%
  group_by(month) %>%
  summarize(
    mean_wbgt = mean(mean_wbgt)
  )

ggplot(monthly, aes(month, mean_wbgt)) + geom_line()

# another one split by year or groups of years
ggplot(monthly_stats, aes(x = month, y = mean_wbgt, group = year, color = factor(year))) +
  geom_line() + labs(x = "Month", y = "Mean WBGT", color = "Year") + theme_minimal()
```

extreme WBGT
```{r}
# percent of California area that has WBGT > 32
ggplot(annual, aes(year, pct_over_32)) + geom_line() + 
  labs(y = "percent of California area that has WBGT > 32")
```


split data into northern and southern CA (split at 35.8 degrees north)
```{r}
# boundary latitude
split_lat <- 35.8

# create north/south extents
south_ext <- ext(-125, -113, 32, split_lat)
north_ext <- ext(-125, -113, split_lat, 43)

# crop CA polygon into two regions
ca_south <- crop(ca_shape, south_ext)
ca_north <- crop(ca_shape, north_ext)

# function to calculate stats
calc_stats <- function(file, region_shape, region_name) {

  r <- rast(file)

  # project polygon if needed
  region_shape <- project(region_shape, crs(r))

  # crop + mask
  r_region <- r |>
    crop(region_shape) |>
    mask(region_shape)

  vals <- values(r_region, na.rm = TRUE)

  tibble(
    region = region_name,
    min_wbgt = min(vals, na.rm = TRUE),
    mean_wbgt = mean(vals, na.rm = TRUE),
    max_wbgt = max(vals, na.rm = TRUE),
    pct_over_32 = mean(vals > 32, na.rm = TRUE)
  )
}

# loop through files
regional_stats <- map2_dfr(
  df_ca$file,
  seq_len(nrow(df_ca)),
  \(f, i) {

    bind_rows(
      calc_stats(f, ca_north, "Northern CA"),
      calc_stats(f, ca_south, "Southern CA")
    ) |>
      mutate(
        year = df_ca$year[i],
        month = df_ca$month[i]
      )
  }
)
```


time series split by northern and southern CA
```{r}
annual_region <- regional_stats %>%
  group_by(region, year) %>%
  summarize(
    min_wbgt = min(min_wbgt),
    mean_wbgt = mean(mean_wbgt),
    max_wbgt = max(max_wbgt),
    pct_over_32 = mean(pct_over_32),
    .groups = "drop"
  )

ggplot(annual_region, aes(x = year, y = mean_wbgt, color = region)) + geom_line() +
  labs(x = "Year", y = "Mean WBGT", color = "Region") + theme_minimal()
```


Annual mean WBGT trend map
```{r}
years <- str_extract(basename(files_ca), "\\d{4}") |> as.integer()
months <- str_extract(
  basename(files_ca),
  "(?<=_)\\d{2}(?=_CA\\.nc)"
) |> as.integer()


# annual mean WBGT raster
unique_years <- sort(unique(years))

annual_mean_list <- list()

for(y in unique_years){

  yr_files <- files_ca[years == y]

  r <- rast(yr_files)

  # annual mean across all timesteps. Mean WBGT for that year at every grid cell.
  annual_mean <- app(r, mean, na.rm=TRUE)

  annual_mean_list[[as.character(y)]] <- annual_mean
}

# stack annual rasters
annual_stack <- rast(annual_mean_list)  # Build a time series per pixel
names(annual_stack) <- unique_years

# compute trend per gridcell
trend_fun <- function(x){

  if(all(is.na(x))) return(NA)

  yrs <- unique_years

  fit <- lm(x ~ yrs) # Fit a linear regression at each grid cell

  slope <- coef(fit)[2]

  # convert to °C/year
  slope
}

trend_map <- app(
  annual_stack,
  trend_fun
)

# plot trend map. Each pixel contains change in annual mean WBGT per year
# Example interpretation: +0.35 °C/year means: Annual mean WBGT increased by 0.35°C every 10 years at that location.

plot(trend_map, main = "Annual Mean WBGT Trend (°C/year)")
```


trend test: Mann-Kendall test and Theil-Sen slope
```{r}
library(rkt)

files_ca <- list.files("wbgt_ca", pattern = "\\.nc$", full.names = TRUE)
df_ca <- data.frame(file = files_ca)
df_ca$year <- str_extract(basename(files_ca), "\\d{4}") |> as.integer()
df_ca$month <- str_extract(basename(files_ca), "(?<=_)\\d{2}(?=_CA\\.nc)") |> as.integer()



# state-wide mean per month
df_ca$wbgt_mean <- map_dbl(df_ca$file, \(f) {
  r <- rast(f)
  global(r, mean, na.rm = TRUE)[1,1]
})

# create time variable. Note that "15" is used as an arbitrary day place-holder
df_ca$date <- as.Date(paste(df_ca$year, df_ca$month, "15", sep = "-"))


# seasonal MK trend test
mk_seasonal <- rkt(
  df_ca$year,
  df_ca$wbgt_mean,  # state-wide mean of each month
  block = df_ca$month)

# results
print(mk_seasonal)

# Theil-Sen slope
mk_seasonal$B

# plot trend
fit <- lm(wbgt_mean ~ date, data = df_ca)
plot(df_ca$date, df_ca$wbgt_mean, pch = 16, xlab="date (monthly)", ylab="state-wide WBGT mean")
abline(fit, col = "red", lwd = 2)
```

mean WBGT map, shows WBGT hotspots
```{r}
overall_mean <- app(annual_stack, mean)
plot(overall_mean, main="Mean WBGT")
# Each grid cell represents the mean WBGT (of all years) at that location
```


ambient temp spatial map -- use Tair from input files (represents t2m in Celsius)
```{r}
# mean ambient temp over all time steps

```




Mann-Kendall test in several locations
```{r}
regions <- list(
  LA = ext(-119, -117, 33.5, 34.5),
  SoCAB = ext(-118.5, -116, 32.5, 34.5),
  CentralValley = ext(-122.5, -119, 35, 39.5),
  ImperialValley = ext(-116.5, -114.5, 32.5, 33.5)
)

regional_mean <- function(file, region_ext){
  r <- rast(file)
  r_crop <- crop(r, region_ext) # crop to region
  global(r_crop, mean, na.rm=TRUE)[1,1] # regional mean
}

# monthly time series for each region
region_results <- list()

for(region_name in names(regions)){
  cat("Processing:", region_name, "\n")
  
  vals <- map_dbl(
    df_ca$file,
    \(f) regional_mean(f, regions[[region_name]])
  )

  tmp <- df_ca |>
    select(file, year, month) |>
    mutate(
      region = region_name,
      wbgt_mean = vals
    )

  region_results[[region_name]] <- tmp
}

regional_df <- bind_rows(region_results)



############### seasonal Mann-Kendall by region
library(rkt)

mk_results <- regional_df |>
  group_by(region) |>
  group_modify(\(d, g){

    mk <- rkt(
      d$year,
      d$wbgt_mean,
      block = d$month
    )

    data.frame(
      tau = mk$tau,
      pval = mk$sl,
      slope = mk$B
    )
  })

print(mk_results)


# plot regional time series
regional_df$date <- as.Date(paste(regional_df$year, regional_df$month, "15", sep = "-"))

ggplot(regional_df,
       aes(date, wbgt_mean, color=region)) +

  geom_line(alpha=0.6) +

  geom_smooth(
    method="lm",
    se=FALSE,
    linewidth=1.2
  ) +

  labs(
    title="Monthly Mean WBGT Trends by Region",
    x="Year",
    y="Mean WBGT (°C)"
  ) +

  theme_minimal()
```


Mann-Kendall on annual statewide WBGT
```{r}
annual_ca <- monthly_stats |>
  group_by(year) |>
  summarize(mean_wbgt = mean(mean_wbgt, na.rm=TRUE))

# MK test
library(Kendall)

mk_annual <- MannKendall(annual_ca$mean_wbgt)
print(mk_annual)

# Sen slope
library(trend)

sen_annual <- sens.slope(annual_ca$mean_wbgt)
print(sen_annual)

# plot annual trend
ggplot(annual_ca,
       aes(year, mean_wbgt)) +

  geom_line() +

  geom_smooth(
    method="lm",
    se=FALSE,
    color="red"
  ) +

  labs(
    title="Annual Statewide Mean WBGT",
    x="Year",
    y="WBGT (°C)"
  ) +

  theme_minimal()
```


Seasonal MK on warm months only
```{r}
warm_df <- monthly_stats |>
  filter(month %in% 5:9)

warm_df$date <- as.Date(paste(warm_df$year, warm_df$month, "15", sep = "-"))

mk_warm <- rkt(
  warm_df$year,
  warm_df$mean_wbgt,
  block = warm_df$month
)

print(mk_warm)

ggplot(warm_df, aes(date, mean_wbgt)) +
  geom_line(alpha=0.7) +
  geom_smooth(method="lm", se=FALSE, color="red") +
  labs(title="Warm-Season Statewide WBGT", x="Year", y="WBGT (°C)") + theme_minimal()
```




