---
title: "calculate_cza_and_fdir"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, results='asis', message=FALSE, warning=FALSE, collapse=TRUE)
```

```{r}
library(heatmetrics)
ls("package:heatmetrics") # functions in the heatmetrics package
```

Data source:

Muñoz Sabater, J. (2019): ERA5-Land hourly data from 1950 to present. Copernicus Climate Change Service (C3S) Climate Data Store (CDS). DOI: 10.24381/cds.e2161bac (Accessed on 27-01-2026)

Hersbach, H., Bell, B., Berrisford, P., Biavati, G., Horányi, A., Muñoz Sabater, J., Nicolas, J., Peubey, C., Radu, R., Rozum, I., Schepers, D., Simmons, A., Soci, C., Dee, D., Thépaut, J-N. (2023): ERA5 hourly data on single levels from 1940 to present. Copernicus Climate Change Service (C3S) Climate Data Store (CDS), DOI: 10.24381/cds.adbb2d47 (Accessed on 27-01-2026)

```{r}
library(ncdf4)
library(terra)
```

```{r}
# open and view preprocessed dataset
wbgt_inputs_aug_2019_new <- nc_open("wbgt_inputs_aug_2019_new.nc", write = TRUE)
names(wbgt_inputs_aug_2019_new$var)
dim(ncvar_get(wbgt_inputs_aug_2019_new, "solar"))
# R shows dimensions as longitude x latitude x times
```

## Calculating "cza"

```{r}
# get inputs for calc_cza_int
lat <- ncvar_get(wbgt_inputs_aug_2019_new, "latitude")
lon <- ncvar_get(wbgt_inputs_aug_2019_new, "longitude")
year  <- ncvar_get(wbgt_inputs_aug_2019_new, "year")
month <- ncvar_get(wbgt_inputs_aug_2019_new, "month")
day   <- ncvar_get(wbgt_inputs_aug_2019_new, "day")
hour  <- ncvar_get(wbgt_inputs_aug_2019_new, "hour")

# expand to full grid
grid <- expand.grid(lon = lon, lat = lat, t = seq_along(year))
grid$y   <- year[grid$t]
grid$mon <- month[grid$t]
grid$d   <- day[grid$t]
grid$hr  <- hour[grid$t]

# call calc_cza_int() ---- takes about 10 minutes
cza <- calc_cza_int(grid$lat, grid$lon, grid$y, grid$mon, grid$d, grid$hr)

# reshape back to 3-dimensional array
ntime <- length(year)
nlat <- length(lat)
nlon <- length(lon)

cza_array <- array(cza, dim = c(nlon, nlat, ntime))
```

```{r}
# check that dimensions of cza_array match dimensions of wbgt_inputs_aug_2019
dim(ncvar_get(wbgt_inputs_aug_2019_new, "solar"))
dim(cza_array)  # longitude x latitude x time

summary(cza_array) # check that all values are in [0,1]
```

```{r}
# crop cza_array to California shape

cza_array_for_terra <- aperm(cza_array, c(2,1,3)) # latitude x longitude x time ordering for terra
r_cza <- rast(cza_array_for_terra) # convert array to raster
ext(r_cza) <- c(min(lon), max(lon), min(lat), max(lat))
crs(r_cza) <- "EPSG:4326"
ca <- vect("urban 2019/ca_state/CA_State.shp")
ca <- project(ca, crs(r_cza))
r_cza_ca <- crop(r_cza, ca, snap="out") # crop with whole grid cells on CA boundary
r_cza_ca <- mask(r_cza_ca, ca)

# convert raster in the shape of CA back to array
cza_array_ca <- as.array(r_cza_ca)
cza_array_ca <- aperm(cza_array_ca, c(2,1,3)) # reorder back to longitude x latitude x time

dim(cza_array_ca) # check dimensions of cropped array
```

```{r}
# check that cza_array_ca is actually cropped to CA by plotting it

times <- as.POSIXct(ncvar_get(wbgt_inputs_aug_2019_new, "valid_time"), origin = "1970-01-01", tz = "UTC")
t_index <- which(times == as.POSIXct("2019-08-01 18:00", tz="UTC"))
r_cza_ca <- rast(aperm(cza_array_ca, c(2,1,3)))
ext(r_cza_ca) <- c(min(lon), max(lon), min(lat), max(lat))
crs(r_cza_ca) <- "EPSG:4326"
plot(r_cza_ca[[t_index]])
title(main = paste("cza at", times[t_index]), line=3)
```

```{r}
#### ONLY RUN IF cza IS NOT ALREADY IN wbgt_inputs_aug_2019.nc

# put cza_array_ca into wbgt_inputs_aug_2019 as a variable -- edited wbgt_inputs_aug_2019 directly

nc_close(wbgt_inputs_aug_2019)

wbgt_inputs_aug_2019 <- nc_open("wbgt_inputs_aug_2019.nc", write = TRUE)

dim_lon  <- wbgt_inputs_aug_2019$dim$longitude
dim_lat  <- wbgt_inputs_aug_2019$dim$latitude
dim_time <- wbgt_inputs_aug_2019$dim$valid_time

cza <- ncvar_def(
  name = "cza",
  units = "",   # no units because it's a ratio
  dim = list(dim_lon, dim_lat, dim_time),
  missval = -9999,
  longname = "cosine solar zenith angle (0-1)",
  prec = "float"
)

wbgt_inputs_aug_2019 <- ncvar_add(wbgt_inputs_aug_2019, cza)

ncvar_put(wbgt_inputs_aug_2019, cza, cza_array_ca)

nc_close(wbgt_inputs_aug_2019)
```

```{r}
# check wbgt_inputs_aug_2019_new

wbgt_inputs_aug_2019_new <- nc_open("wbgt_inputs_aug_2019_new.nc")
names(wbgt_inputs_aug_2019_new$var)
dim(ncvar_get(wbgt_inputs_aug_2019_new, "cza"))
# R shows dimensions as longitude x latitude x time
```

```{r}
# just checking
# plot with wbgt_inputs_aug_2019 cza variable at a single valid_time

r <- rast("wbgt_inputs_aug_2019_new.nc", subds = "cza")
time_raw <- ncvar_get(wbgt_inputs_aug_2019_new, "valid_time")
time_units <- ncatt_get(wbgt_inputs_aug_2019_new, "valid_time", "units")$value
time_vals <- as.POSIXct(time_raw, origin="1970-01-01", tz="UTC")
time(r) <- time_vals
t_index <- which(time_vals == as.POSIXct("2019-08-01 18:00", tz="UTC"))
plot(r[[t_index]])
title(main = "cza at 2019-08-01 18:00", line=3)
```

## Calculating "fdir"

```{r}
# get inputs for calc_fdir
lat <- ncvar_get(wbgt_inputs_aug_2019_new, "latitude")
lon <- ncvar_get(wbgt_inputs_aug_2019_new, "longitude")
year  <- ncvar_get(wbgt_inputs_aug_2019_new, "year")
month <- ncvar_get(wbgt_inputs_aug_2019_new, "month")
dday   <- ncvar_get(wbgt_inputs_aug_2019_new, "dday")  # needs to be day-fractions
solar <- ncvar_get(wbgt_inputs_aug_2019_new, "fdir_W")
cza <- ncvar_get(wbgt_inputs_aug_2019_new, "cza")

# expand to full grid
grid <- expand.grid(lon = lon, lat = lat, t = seq_along(year))
grid$y   <- year[grid$t]
grid$mon <- month[grid$t]
grid$d   <- dday[grid$t]
grid$solar <- solar[grid$t]
grid$cza <- cza[grid$t]

# create valid-cell mask
valid <- !is.nan(grid$solar) & !is.nan(grid$cza)

fdir <- rep(NA, nrow(grid))

# call calc_fdir() ---- takes about 5 minutes
fdir[valid] <- mapply(calc_fdir, grid$y[valid], grid$mon[valid], grid$d[valid], grid$lat[valid],
                         grid$lon[valid], grid$solar[valid], grid$cza[valid])

# reshape back to 3-dimensional array
ntime <- length(year)
nlat <- length(lat)
nlon <- length(lon)

fdir_array <- array(fdir, dim = c(nlon, nlat, ntime))
```

```{r}
# check that dimensions of fdir_array match dimensions of wbgt_inputs_aug_2019_new
dim(ncvar_get(wbgt_inputs_aug_2019_new, "solar"))
dim(fdir_array)  # longitude x latitude x time

summary(fdir_array) # check that all values are in [0,1]

length(unique(fdir_array[!is.na(fdir_array)])) # checking that there are different values
```
```{r}
# crop fdir_array to California shape

fdir_array_for_terra <- aperm(fdir_array, c(2,1,3)) 
# latitude x longitude x time ordering for terra
r_fdir <- rast(fdir_array_for_terra) # convert array to raster
ext(r_fdir) <- c(min(lon), max(lon), min(lat), max(lat))
crs(r_fdir) <- "EPSG:4326"
ca <- vect("urban 2019/ca_state/CA_State.shp")
ca <- project(ca, crs(r_fdir))
r_fdir_ca <- crop(r_fdir, ca, snap="out") # crop with whole grid cells on CA boundary
r_fdir_ca <- mask(r_fdir_ca, ca)

# convert raster in the shape of CA back to array
fdir_array_ca <- as.array(r_fdir_ca)
fdir_array_ca <- aperm(fdir_array_ca, c(2,1,3)) # reorder back to longitude x latitude x time

dim(fdir_array_ca) # check dimensions of cropped array
```

```{r}
# check that fdir_array_ca is actually cropped to CA by plotting it

times <- as.POSIXct(ncvar_get(wbgt_inputs_aug_2019_new, "valid_time"), origin = "1970-01-01", tz = "UTC")
t_index <- which(times == as.POSIXct("2019-08-01 18:00", tz="UTC"))
r_fdir_ca <- rast(aperm(fdir_array_ca, c(2,1,3)))
ext(r_fdir_ca) <- c(min(lon), max(lon), min(lat), max(lat))
crs(r_fdir_ca) <- "EPSG:4326"
plot(r_fdir_ca[[t_index]])
title(main = paste("fdir at", times[t_index]), line=3)
```

```{r}
####### NEED TO FIX SAVING TO THE FILE--- GOT WEIRD "" NAMES --- JUST CREATE A NEW FILE

# make a copy of wbgt_inputs_aug_2019 and then add fdir to it

```

```{r}
# check wbgt_inputs_aug_2019_new

wbgt_inputs_aug_2019 <- nc_open("wbgt_inputs_aug_2019_new.nc")
names(wbgt_inputs_aug_2019_new$var)
dim(ncvar_get(wbgt_inputs_aug_2019_new, "cza"))
# R shows dimensions as longitude x latitude x time
```

## Calculating WBGT "fdir" input using calc_solar_parameters

```{r}
# get inputs for calc_solar_parameters
lat <- ncvar_get(wbgt_inputs_aug_2019_new, "latitude")
lon <- ncvar_get(wbgt_inputs_aug_2019_new, "longitude")
year  <- ncvar_get(wbgt_inputs_aug_2019_new, "year")
month <- ncvar_get(wbgt_inputs_aug_2019_new, "month")
day   <- ncvar_get(wbgt_inputs_aug_2019_new, "day")
solar <- ncvar_get(wbgt_inputs_aug_2019_new, "solar")  # ERA5-Land ssrd in W/m2
cza <- ncvar_get(wbgt_inputs_aug_2019_new, "cza")  # calculated from calc_cza_int()
fdir <- ncvar_get(wbgt_inputs_aug_2019_new, "fdir_frac") # (NN-inter ERA5 fdir)/(ERA5-Land ssrd)

# expand to full grid
grid <- expand.grid(lon = lon, lat = lat, t = seq_along(year))
grid$y   <- year[grid$t]
grid$mon <- month[grid$t]
grid$d   <- day[grid$t]
grid$solar <- solar[grid$t]
grid$cza <- cza[grid$t]
grid$fdir <- fdir[grid$t]

# call calc_solar_parameters() ---- takes about 10 minutes
solar_params <- mapply(calc_solar_parameters,grid$y, grid$mon, grid$d,grid$lat, grid$lon, grid$solar, grid$cza, grid$fdir)

# reshape back to 3-dimensional array
ntime <- length(year)
nlat <- length(lat)
nlon <- length(lon)

solarRet_array <- array(as.numeric(unlist(solar_params["solarRet", ])), dim = c(nlon, nlat, ntime))
cza_array      <- array(as.numeric(unlist(solar_params["cza", ])), dim = c(nlon, nlat, ntime))
fdir_array     <- array(as.numeric(unlist(solar_params["fdir", ])), dim = c(nlon, nlat, ntime))
```

```{r}
# check that dimensions of fdir_array match dimensions of wbgt_inputs_aug_2019_new
dim(ncvar_get(wbgt_inputs_aug_2019_new, "solar"))
dim(fdir_array)  # longitude x latitude x time

length(unique(fdir_array[!is.na(fdir_array)]))  # checking that there are different values

summary(solarRet_array)
summary(fdir_array)
```

```{r}
# -----------crop solarRet_array to California shape

solarRet_array_for_terra <- aperm(solarRet_array, c(2,1,3)) 
# latitude x longitude x time ordering for terra
r_solarRet <- rast(solarRet_array_for_terra) # convert array to raster
ext(r_solarRet) <- c(min(lon), max(lon), min(lat), max(lat))
crs(r_solarRet) <- "EPSG:4326"
ca <- vect("urban 2019/ca_state/CA_State.shp")
ca <- project(ca, crs(r_solarRet))
r_solarRet_ca <- crop(r_solarRet, ca, snap="out")   # crop to bounding box
r_solarRet_ca <- mask(r_solarRet_ca, ca)           # mask outside CA to NA


# convert raster in the shape of CA back to array
solarRet_array_ca <- as.array(r_solarRet_ca)
solarRet_array_ca <- aperm(solarRet_array_ca, c(2,1,3)) # reorder back to longitude x latitude x time

dim(solarRet_array_ca) # check dimensions of cropped array



# -------------crop fdir_array to California shape

fdir_array_for_terra <- aperm(fdir_array, c(2,1,3)) 
# latitude x longitude x time ordering for terra
r_fdir <- rast(fdir_array_for_terra) # convert array to raster
ext(r_fdir) <- c(min(lon), max(lon), min(lat), max(lat))
crs(r_fdir) <- "EPSG:4326"
ca <- vect("urban 2019/ca_state/CA_State.shp")
ca <- project(ca, crs(r_fdir))
r_fdir_ca <- crop(r_fdir, ca, snap="out") # crop with whole grid cells on CA boundary
r_fdir_ca <- mask(r_fdir_ca, ca)

# convert raster in the shape of CA back to array
fdir_array_ca <- as.array(r_fdir_ca)
fdir_array_ca <- aperm(fdir_array_ca, c(2,1,3)) # reorder back to longitude x latitude x time

dim(fdir_array_ca) # check dimensions of cropped array
```

```{r}
# check that solarRet_array_ca is actually cropped to CA by plotting it

times <- as.POSIXct(ncvar_get(wbgt_inputs_aug_2019_new, "valid_time"), origin = "1970-01-01", tz = "UTC")
t_index <- which(times == as.POSIXct("2019-08-01 18:00", tz="UTC"))
r_solarRet_ca <- rast(aperm(solarRet_array_ca, c(2,1,3)))
ext(r_solarRet_ca) <- c(min(lon), max(lon), min(lat), max(lat))
crs(r_solarRet_ca) <- "EPSG:4326"
plot(r_solarRet_ca[[t_index]])
title(main = paste("solarRet at", times[t_index]), line=3)


# check that fdir_array_ca is actually cropped to CA by plotting it

times <- as.POSIXct(ncvar_get(wbgt_inputs_aug_2019_new, "valid_time"), origin = "1970-01-01", tz = "UTC")
t_index <- which(times == as.POSIXct("2019-08-01 18:00", tz="UTC"))
r_fdir_ca <- rast(aperm(fdir_array_ca, c(2,1,3)))
ext(r_fdir_ca) <- c(min(lon), max(lon), min(lat), max(lat))
crs(r_fdir_ca) <- "EPSG:4326"
plot(r_fdir_ca[[t_index]])
title(main = paste("fdir at", times[t_index]), line=3)
```




```{r}
# add solarRet_array_ca and fdir_array_ca to new dataset called wbgt_inputs_aug_2019_new2
## === FIX

library(ncdf4)

nc_in <- nc_open("wbgt_inputs_aug_2019_new.nc")

# define dimensions
lon_dim <- nc_in$dim$longitude
lat_dim <- nc_in$dim$latitude
time_dim <- nc_in$dim$valid_time


# Define new variables
solarRet <- ncvar_def(name="solarRet", units = "W m**-2", 
                      dim = list(lon_dim, lat_dim, time_dim), 
                      missval = -9999, longname = "adjusted solar radiation", prec="float")
fdir <- ncvar_def(name="fdir", units = "", 
                          dim = list(lon_dim, lat_dim, time_dim), 
                          missval = -9999, longname = "fraction of irradiance due to direct beam",
                          prec = "float")

# Create new NetCDF file
nc_out <- nc_create("wbgt_inputs_aug_2019_new2.nc", vars = list(solarRet, fdir), force_v4 = TRUE)


# Copy existing data
for (varname in names(nc_in$var)) {
  data <- ncvar_get(nc_in, varname)
  old_var <- nc_in$var[[varname]]
  
  new_var <- ncvar_def(
    name = old_var$name,
    units = old_var$units,
    dim = lapply(old_var$dim, function(d) d),
    missval = ifelse(is.null(old_var$missval), -9999, old_var$missval),
    longname = old_var$longname,
    prec = old_var$prec
  )
  
  # add variable to NetCDF
  ncvar_obj <- ncvar_add(nc_out, new_var)
  
  # write data
  ncvar_put(nc_out, ncvar_obj, data)
}

# Write new arrays
ncvar_put(nc_out, "solarRet", solarRet_array_ca)
ncvar_put(nc_out, "fdir",     fdir_array_ca)

# Close files
nc_close(nc_in)
nc_close(nc_out)




```

